Lens Modeling in \sw

### 2014-06-17 text/partA.tex
editing

The first discoveries of gravitational lenses, in each case a galaxy
lensing a background quasar into multiple images
\citep{1979Natur.279..381W,1980Natur.285..641W}, brought in their wake
the first work on lens modelling
\citep{1981ApJ...244..723Y,1981ApJ...244..736Y}.  For one of the
lenses, mass modelling predicted that one of the lensed quasar images
seen would split further into two at higher resolution, which was
crucial in subsequently confirming that the object is indeed an
example of lensing.  It had long been expected that galaxies must
sometimes cause lensed images \citep{1937ApJ....86..217Z}, and it even
been argued that the phenomenon could help measure cosmological
parameters \citep{1964MNRAS.128..307R,1966MNRAS.132..101R}.  But
apparently nobody was expecting that lenses would need detailed
modelling.  In hindsight, the reason is not difficult to understand.
The scale of image separations is set by the angular Einstein radius
\begin{equation}
\sim \left(\frac{4GM}{c^2 D_L}\right)^{1/2}
\simeq 0.1'' \left(\frac{M}{M_\odot}\right)^{1/2}
             \left(\frac{D_L}{\rm pc}\right)^{-1/2}
\end{equation}
where $D_L$ is the distance to the lens, and $M$ its mass.  For
stellar-mass lenses, the image separation is much larger than the size
of the lensing mass, so lenses are well approximated by point masses.
In galaxy lenses, the image separation is comparable to the size of
the galaxy; typically the lensed images are seen through the galaxy
halo.  Hence the lensed images depend on the detailed mass
distribution of the lensing galaxy.  Once observations of the lensed
images became available, it was obvious that mass modelling was
needed.

Since those first discoveries, more than 400 secure lenses are now
known (see Figure \ref{fig:masterlens}).  Modelling of the mass
distribution is part of any research using lenses, but so far no
modelling study has spanned all known lenses.  The largest single one
\citep{2009ApJ...703L..51K} models 58 lenses to infer the distribution
of dark matter around galaxies.  In other work,
\cite{2011ApJ...740...97L} combined lens models of 21 galaxies with
models of their stellar populations, to find the relation between
stars and dark matter, and \cite{2014MNRAS.437..600S} modelled 18
time-delay lenses together to infer cosmological parameters.

@@

Surveys now under way aim to increase the inventory of lenses another
ten or a hundred fold.  Among these is \sw (Marshall et al, in prep,
More et al, in prep).  \sw is a citizen science project\footnote{\tt
  http://www.spacewarps.org} in which volunteers are presented
sky-survey images and invited to identify lens candidates.  Simulated
lenses are mixed in with the data, both to help train volunteers on
what to look for, and to provide measures of the effectiveness of the
search.  The motivation for \sw is to enable volunteers, some of whom
had previously serendipitously identified lens candidates on earlier
citizen-science surveys, to make discoveries missed in automatic
searches by software robots.  Robots are very good with clean lensing
system in uncrowded fields with high signal-to-noise, but in general
test situations \citep{2009ApJ...694..924M} robots miss lenses (low
completeness) or contaminate the results with non-lenses (low purity).
\sw introduced its first tranche of survey data, in May 2013.  The
Canada-France-Hawaii Telescope Legacy Survey (CFHTLS), covering
$\simeq172$ square degrees (0.4\% of the sky) was divided into tiles
of $80''\times80''$ ($440\times 440$ pixels).  Each such tile was
shown to ten volunteers.  Despite the rarity of detectable lenses
($<10^{-3}$ per tile), the \sw collective soon found lens candidates,
both rediscoveries already known from robots \citep{Gavazzi2012,
  More2012ApJ} and possible new lenses.

The encouraging early results prompted the question: could modeling of
the lenses also be done by volunteers?  An early proposal for what
became \sw included a prototype lens modelling tool
\citep{2010AAS...21543527N}.  There are several software tools for
lensing modelling available, and work has been done on generic
interfaces \citep{2013NewAR..57....1L,2014A&C.....5...28L}.  Moreover,
some {\em Spacewarps\/} volunteers are quite experienced from earlier
projects, having spent a thousand hours or more with data, and very
interested in more demanding projects.  The interests of
citizen-science communities are just beginning to be studied
\citep[e.g.,][]{2013AEdRv..12a0106J}, but it is clear that some
volunteers welcome open-ended challenges, and sometimes these have led
to new scientific results: one example is the discovery of an
exceptional extra-solar planet \citep{2013ApJ...768..127S}; another is
the development of new algorithms for protein folding
\citep{Khatib22112011}.  All these are grounds for optimism.  There
is, however, a basic difficulty in strong gravitational lensing.
Lensed images do not look much like their source, and still less do
they resemble the lensing-mass distribution.  To model a lens, one
needs to either do a lot of random guessing, or have a good intuition
for what works.

In this paper, we propose a way around the difficulty, and report on a
modelling test on \sw using simulated lenses.  There are three aspects
to this work, covered in the following three sections.

In section \secref{Fermat} we introduces a theoretical innovation,
which we call a ``spaghetti diagram''.  This is a cartoon-like markup
of a lensed system, which implicitly encodes the basis of a mass
model.  A spaghetti diagram is basically the saddle-point contours
originally introduced to gravitational lensing by
\cite{1986ApJ...310..568B} as a way of classifying lensed images.
They are sometimes shown as part of the output of lens models
\citep[for
  example][]{2001ApJ...557..594R,2003ApJ...590...39K,Lubini2012}.  In
the present work, however, spaghetti diagrams are the {\em input\/}
through which the modeller tells the program what to do.

In section \secref{SpaghettiLens} we describe the \spl program, which
implements everything.  \spl is an extension of the GLASS framework
for modelling lenses \citep{2014arXiv1401.7990C}.  \spl is designed to
be friendly to the forum style of citizen-science projects, without
sacrificing any features of GLASS.  The program is invoked online, and
the modelling history is logged in a transparent way, enabling forum
discussions and incremental improvement by different people.
Moreover, any result can be revisited and post-processed if desired.
These features are of general interest for citizen-science projects,
but somewhat outside the scope of the present paper.

The purpose of this study was to provide a means to model a large number of gravitational lenses by showing that gravitational lens modeling can be learned / done by volunteers.
We suggest, that volunteers will be as successful as professionals with modeling if provided with an easy to use tool with visual feedback (What you see is what you get, WYSIWYG) and a minimal set of instructions.
Volunteers will then crowd work (using buzz word here ;) ) / work collaborative on modeling lenses from several sources / groups at a central place.
Since this is an iterative learning process, the more involved volunteers will quickly gain knowledge that can be passed down to new volunteers.
That creates a social structure that scales well with the number of volunteers, as other projects have already shown. %TODO \needcite
Finding people working as volunteers has been shown to be successful last but not least by \sw and the whole galaxy zoo project.
To test the people's abilities, we investigated the performance of a few volunteers modeling a set of simulated lenses.
We tested the ability to correctly identify lensed images and reproduce similar mass map of the lens.

It is well known that making a scientific contribution is a central
motivation cited by citizen-science volunteers.  So for a lens
modeler in a citizen-science project, nobody wants that it should
sacrifice scientific usefulness in order to be fun to use.  That said,
being aesthetically pleasing and having a short initial learning curve
are also essential qualities.  Also desirable is that the user is
encouraged/challenged to go deeper and wider into the subject.
  \spl tries to address all of these.

@@

We now explain spaghetti diagrams and then outline what \spl does with
them.

@@

Let us recall the formulation of gravitational lensing by
\cite{1986ApJ...310..568B} in terms of Fermat's principle .  Consider
a lens at some redshift $z_L$ and let $(x,y)$ be planar coordinates at
the lens, transverse to the line of sight.  Let $\Sigma(x,y)$ be the
mass per unit area, projecting along the line of sight.  This surface
density is often given in a dimensional form
\begin{equation} \label{eq:kappa}
\kappa(x,y) \propto \Sigma(x,y) \,.
\end{equation}
Consider also a more distant light source, at redshift $z_S$, behind
point $(x_s,y_s)$ on the lens.

Now imagine a virtual photon coming from the source to some $(x,y)$ on
the lens, then changing direction and coming to the observer.  Such a
direction change would increase the light travel time compared to
coming through $(x_s,y_s)$.  The increased light travel time from the
geometry of deflection would be
\begin{equation} \label{eq:tgeom}
\tgeom(x,y) \propto (x-x_s)^2 + (y-s_y)^2 \,.
\end{equation}
assuming it is small compared to the total light travel time.

@@

The proportionality factors in \eqref{eq:kappa} and \eqref{eq:tgeom}
depend on the redshifts and cosmological parameters and are given in
the Appendix~\ref{more-theory}.

@@

The full function $t(x,y)$, also known as the arrival-time surface,
applies to virtual photons.  In other words, it is an abstract
construct and not itself observable.  But observable image positions
can be derived from the arrival-time surface, so visualising it is
useful.  Figure~\figref{arriv} does so.  In this figure, a maximum,
where present, is easy to see.  To locate mimima and saddle-points,
however, one needs to examine the contours of equal arrival time.  A
saddle point is characterised by a contour crossing itself, forming
an~X.  Mimima, on the other hand, have contours looping around them,
as do maxima.

The saddle-point contours which form an X are especially interesting,
because they set the overall topography of the arrival-time surface.
The obviously give the locations of the saddle points, and roughly
localise the minima and maxima as well.  If more precise locations of
the minima and maxima are added, the whole arrival-time surface is
already approximately known.  Since the arrival-time surface has an
exact relation to the lens-mass distribution and the source position,
in effect the mass distribution is also automatically specified.  In
other words, a simple sketch of saddle-point contours along with
locations of minima and maxima ---which we call a spaghetti diagram---
is implicitly already an approximation to a lens-mass distribution.
This idea is the basis of \spl.

@@

The first modeling step with \spl is for the user to call up the \sw
of interest, and make an educated guess for the topography of the
arrival-time surface --- specifically, the locations of the maxima,
minima and saddle points that correspond to images of the brightest
part of the source.  Input is given by tracing a proposal for the
saddle-point contours.  \Figref{input-spag} shows some example
inputs --- and incidentally also indicates the origin of the name
\spl.

@@

The exact placement of the pink spaghetti loops shown in
\Figref{input-spag} has no significance.  Their utility is
just to help mark up the image as maximum, minima and saddle points.
As the figure suggests, the marking up can be done very easily with a
mouse.  Human intuition is required to (a)~identify lensed images and
separate them from other background light sources, and (b)~classify
and order images according to arrival time.  But the computer helps by
allowing only valid lensing configurations to be entered, and ensuring
that the odd image theorem is taken care of.%TODO \needcite

>>>>

The first two discoveries of gravitational lenses
\citep{1979Natur.279..381W,1980Natur.285..641W}, where a galaxy images
a background quasar into two (four), brought in their wake the first
work on lens modelling
\citep{1981ApJ...244..723Y,1981ApJ...244..736Y}.  For the second lens
(PG1115+080), mass models scored an early success with the prediction
that one of the lensed images seen would split further into two at
higher resolution.  That galaxies must sometimes cause had long been
expected \citep{1937ApJ....86..217Z}, and it had even been argued that
the phenomenon could help measure cosmological parameters
\citep{1964MNRAS.128..307R,1966MNRAS.132..101R}.  But apparently
nobody was expecting that lenses would need detailed modelling.
The first observations, however, immediately stimulated models.
The reason for that lies in the image separation. Recall that image
separations are of order the angular Einstein radius
\begin{equation}
\sim \left(\frac{4GM}{c^2 D_L}\right)^{1/2}
\simeq 0.1'' \left(\frac{M}{M_\odot}\right)^{1/2}
             \left(\frac{D_L}{\rm pc}\right)^{-1/2}
\end{equation}
where $D_L$ is the distance to the lens, and $M$ its mass.  For
stellar-mass lenses, the image separation is much larger than the size
of the lensing mass, so lenses are well approximated by point masses.
In galaxy lenses, the image separation is comparable to the size of
the galaxy; typically the lensed images are seen through the galaxy
halo.  Hence the lensed images depend on the detailed mass
distribution of the lensing galaxy.  Galaxy lenses demand modelling.

Since the early discoveries, more than 400 secure lenses are now known
(see Figure \ref{fig:masterlens}).  Modelling of the mass distribution
is part of any research using lenses, but so far no modelling study
has spanned all known lenses.  The largest single one
\citep{2009ApJ...703L..51K} models 58 lenses to infer the distribution
of dark matter around galaxies.  In other work,
\cite{2011ApJ...740...97L} combined lens models of 21 galaxies with
models of their stellar populations, to find the relation between
stars and dark matter, and \cite{2014MNRAS.437..600S} modelled 18
time-delay lenses together to infer cosmological parameters.

@@

Surveys now under way aim to increase the inventory of lenses another
ten or a hundred fold.  Among these is \sw (Marshall et al, in prep,
More et al, in prep).  \sw is a citizen science project\footnote{\tt
  http://www.spacewarps.org} in which volunteers are presented
sky-survey images and invited to identify lens candidates.  Simulated
lenses are mixed in with the data, both to help train volunteers on
what to look for, and to provide measures of the effectiveness of the
search.  The motivation for \sw is to enable volunteers, some of whom
had previously serendipitously identified lens candidates on earlier
citizen-science surveys, to make discoveries missed in automatic
searches by software robots.  Robots are very good with clean lensing
system in uncrowded fields with high signal-to-noise, but in general
test situations \citep{2009ApJ...694..924M} robots miss lenses (low
completeness) or contaminate the results with non-lenses (low purity).
\sw introduced its first tranche of survey data, in May 2013, from the
Canada-France-Hawaii Telescope Legacy Survey (CFHTLS), covering
$\simeq172$ square degrees (0.4\% of the sky) was divided into tiles
of $\simeq 80''\times80''$ ($440\times 440$ pixels).  Each such tile
was shown to ten volunteers.  Despite the rarity of detectable lenses
($<10^{-3}$ per tile), the \sw collective soon found lens candidates,
both rediscoveries already known from robots \citep{Gavazzi2012,
  More2012ApJ} and possible new lenses.

The encouraging early results prompted the question: could modeling of
the lenses also be done by volunteers?  There are several software
tools for lensing modelling available, and work has been done on
generic interfaces \citep{2013NewAR..57....1L,2014A&C.....5...28L}.
Some early designs for \sw included a prototype lens modelling tool
\citep{2010AAS...21543527N}.  Moreover, some {\em Spacewarps\/}
volunteers are quite experienced from earlier projects, having spent a
thousand hours or more with data, and are very interested in more
demanding projects.  The interests of citizen-science communities are
just beginning to be studied \citep[e.g.,][]{2013AEdRv..12a0106J}, but
it is clear that some volunteers welcome open-ended challenges, and
sometimes these have led to new scientific results: one example is the
discovery of an exceptional extra-solar planet
\citep{2013ApJ...768..127S}; another is the development of new
algorithms for protein folding \citep{Khatib22112011}.  All these are
grounds for optimism.  There is, however, a basic difficulty in strong
gravitational lensing.  Lensed images do not look much like their
source, and still less do they resemble the lensing-mass distribution.
To model a lens, one needs to either do a lot of random guessing, or
have a good intuition for what works.

In this paper, we propose a way around the difficulty, and report on a
modelling test on \sw using simulated lenses.  The three following
sections are devoted to the concept, the implementation, and tests
respectively.

In \secref{Fermat} we introduce a markup system for lensed images,
which we call a ``spaghetti diagram''.  A spaghetti diagram resembles
the visible image system, in a cartoon-like way, and at the same time
it encodes the basis of a mass model.  This supplies an intuitive link
between the image system and the mass distribution, which look
frustratingly different from each other.  Spaghetti diagrams are
essentially the saddle-point contours originally introduced to
gravitational lensing by \cite{1986ApJ...310..568B} as a way of
classifying lensed images.  They are sometimes shown as part of the
output of lens models \citep[for
  example][]{2001ApJ...557..594R,2003ApJ...590...39K,Lubini2012}.  In
the present work, however, spaghetti diagrams are the {\em input\/}
through which the modeller tells the program what to do.

In \secref{SpaghettiLens} we describe the \spl program, which
implements everything.  \spl is an extension of the GLASS framework
for modelling lenses \citep{2014arXiv1401.7990C}.  We will not go into
software details in this paper, concentrating instead on lens
modelling per se, but we remark that \spl is designed to be friendly
to the forum style of citizen-science projects, without sacrificing
any features of GLASS.  The program is invoked online, and the
modelling history is logged in a transparent way, enabling forum
discussions and incremental improvement by different people.
Moreover, any result can be revisited and post-processed if desired.

In \secref{mod_challenge} we describe a modelling challenge where a
diverse sample of 29 simulated lenses was modelled multiple times
using \spl.  The models are then examined in two ways.  One was
whether the spaghetti diagram was correct.  The other was the recovery
of the Einstein radius.  In addition, we show some visual comparisons
of actual and recovered lens shape and radial profile, and identify
some areas to improve, but do not compute statistics on these aspects.
Profile and shape recovery with GLASS has been studied in more detail
in \citep{2014arXiv1401.7990C}.

\secref{todo} concludes, with the general outlook and next steps.

@@

@@

Let us recall the formulation of gravitational lensing by
\cite{1986ApJ...310..568B} in terms of Fermat's principle.  Consider a
lens at some redshift $z_L$ and let $(x,y)$ be planar coordinates at
the lens, transverse to the line of sight.  Let $\Sigma(x,y)$ be the
mass per unit area, i.e., density projected along the line of sight.
This surface density is often given in a dimensionless form
\begin{equation} \label{eq:kappa}
\kappa(x,y) \propto \Sigma(x,y)
\end{equation}
called the convergence.  And let there be light, in the form of a more
distant source, at redshift $z_S$, behind point $(x_s,y_s)$ on the
lens.

We now imagine a virtual photon flying from the source to some $(x,y)$
on the lens, then changing direction and coming to the observer.  Such
a direction change would increase the light travel time compared to
coming through $(x_s,y_s)$.  The increased light travel time from the
geometry of deflection would be
\begin{equation} \label{eq:tgeom}
\tgeom(x,y) \propto (x-x_s)^2 + (y-s_y)^2 \,.
\end{equation}
assuming the delay is small compared to the total light travel time.

@@

The proportionality factors in \eqref{eq:kappa} and \eqref{eq:tgeom}
depend on the redshifts and cosmological parameters, and are given in
the Appendix~\ref{more-theory}.

@@

The full function $t(x,y)$, also known as the arrival-time surface,
applies to virtual photons.  In other words, it is an abstract
construct and not itself observable.  But observable image positions
can be derived from the arrival-time surface, so visualising the
surface is useful.  \figref{arriv} does so.  In this figure, a
maximum, if present, is easy to see.  To locate mimima and
saddle-points, however, one needs to examine the contours of equal
arrival time.  A saddle point is characterised by a contour crossing
itself, forming an~X.  Mimima, on the other hand, have contours
looping around them, as do maxima.

The saddle-point contours which form an X are especially interesting,
because they set the overall topography of the arrival-time surface.
The obviously give the locations of the saddle points, and roughly
localise the minima and maxima as well.  If more precise locations of
the minima and maxima are added, the whole arrival-time surface is
already approximately known.  Since the arrival-time surface has an
exact relation to the lens-mass distribution and the source position,
in effect the mass distribution is also automatically specified.  In
other words, a simple sketch of saddle-point contours along with
locations of minima and maxima ---which we call a spaghetti diagram---
is implicitly already an approximation to a lens-mass distribution.

@@

The first modeling step with \spl is for the user to call up the \sw
of interest, and make an educated guess for the topography of the
arrival-time surface --- specifically, the locations of the maxima,
minima and saddle points that correspond to images of the brightest
part of the source.  Input is given by tracing a proposal for the
saddle-point contours.  Examples appear in Figures \ref{fig:6941} to
\ref{fig:7022} below (upper-left panel in each case).

@@

The exact placement of the pink spaghetti loops shown in
\Figref{input-spag} has no significance.  Their utility is
just to help mark up the image as maximum, minima and saddle points.
As the figure suggests, the marking up can be done very easily with a
mouse.  Human intuition is required to (a)~identify lensed images and
separate them from other background light sources, and (b)~classify
and order images according to arrival time.  But the computer helps by
allowing only valid lensing configurations to be entered, and ensuring
that the odd image theorem is taken care of.
